---
title: "Mark-Recapture Models for Simulated Data"
output:
  html_document:
    df_print: paged
---

### Introduction

Mark-recapture models for analyzing simulated capture history data. These models are simplified versions of the models presented in maps_models_notebook. Simplifications are:

1. Single population.
2. No spatial variance component.
3. No residency structure.
4. Constant capture probablity across years.

#### Set working directories and load packages

Only change the main directory. Rest of the directories are automatically created if they don't exist. First line removes all objects from the global environment. Comment it out if you want to keep the current objects.


### 1. Density dependent model

```{r, message=FALSE}

{
sink("dd_model_sim.jags")
cat(
  "model {
    
    # Part 1: Priors
      
      # Part 1.1: Priors for survival model
      
      survival_ad ~ dunif(0,1)
      survival_juv <- (1-survival_ad)/fecundity
      alpha[2] <- logit(survival_ad)
      alpha[1] <- logit(survival_juv)
      beta ~ dt(0,pow(10,-2),1)
      sigma_s ~ dt(0,pow(10,-2),1)T(0,)
      tau_s <- 1/(sigma_s^2)
      
      for (k in 1:npop) {
        for (t in first_pop[k]:(nyear-1)) {
          epsilon[k,t] ~ dnorm(0, tau_s)
        }
      }
      
      
      # Part 1.2: Priors for capture probability model
      
      gamma[1] ~ dt(0,pow(10,-2),1)
      gamma[2] ~ dt(0,pow(10,-2),1)
      
      # Part 1.3: Priors for fecundity model
      
      fecundity ~ dnorm(0,pow(10,-2))T(0,)
      theta <- log(fecundity)
      zeta ~ dt(0,pow(5,-2),1)
      sigma_f ~ dt(0,pow(5,-2),1)T(0,)
      tau_f <- 1/(sigma_f^2)
      
      for (l in 1:nfec) {
        omega[l] ~ dnorm(0, tau_f)
      }

    
    # Part 2: Robust design mark-recapture model
      
    for (i in 1:nind) {
        
      # Part 2.1: Survival process
        
      z[i,first[i]] ~ dbern(1)
      
      for (t in first[i]:(last_pop[pop[i]]-1)) {
        phi[i,t] <- ilogit(alpha[stage[i,t]] + beta*(density[pop[i],t]-1) + epsilon[pop[i],t])
        z[i,t+1] ~ dbern(phi[i,t]*z[i,t])
      }#t
      
      # Part 2.2: Observation
      
      y[i,first[i],first_sub[i]] ~ dbern(1)
      y.pred[i,first[i],first_sub[i]] ~ dbern(1)
      
      for(h in (first_sub[i]+1):nsub) {
        p[i,first[i],h] <- ilogit(gamma[stage[i,first[i]]])
        y[i,first[i],h] ~ dbern(z[i,first[i]]*p[i,first[i],h])
        y.pred[i,first[i],h] ~ dbern(z[i,first[i]]*p[i,first[i],h])
      }#h
      
      for (t in first[i]:(last_pop[pop[i]]-1)) {
        for(h in 1:nsub) {
          p[i,t+1,h] <- ilogit(gamma[stage[i,t+1]])
          y[i,t+1,h] ~ dbern(z[i,t+1]*p[i,t+1,h])
          y.pred[i,t+1,h] ~ dbern(z[i,t+1]*p[i,t+1,h])
        }#h
      }#t
    }#i
      
    
    # Part 3: Population size and Density Estimation
      
      # Part 3.1: Probability of capture per population and year
      
      py_a <- 1 - pow(1-ilogit(gamma[2]),4)
      py_j <- 1 - pow(1-ilogit(gamma[1]),4)
      
      # Part 3.2: Density estimation
      
      for (k in 1:npop) {
        for (t in first_pop[k]:nyear) {
          Nad[k,t] <- (Nobs_ad[k,t]/py_a) + (1-py_a)/py_a
          Njuv1[k,t] <- (Nobs_juv1[k,t]/py_j) + (1-py_j)/py_j
      
          ## Total population size      
          N[k,t] <- Nad[k,t] + Njuv1[k,t]
      
          ## Population Density
          density[k,t] <- N[k,t]/mean(N[k,])
        }#t
      }#k
      
    
    # Part 4: Fecundity model
      
    for (l in 1:nfec) {
      year[l] <- year.index[l]-1
      density.fec[l] <- density[pop.index[l],year[l]]
      Njuv2[l] <- exp(log(Nad[pop.index[l],year.index[l]]) + theta + zeta*(density.fec[l]-1) + omega[l])
      Nobs_juv2[l] ~ dpois(max(py_j*Njuv2[l], 0.00000001))
      Nobs_juv.pred[l] ~ dpois(max(py_j*Njuv2[l], 0.00000001))
    }#l
      
    
    # Part 5: Model Checking
      
      # Part 5.1: Fecundity
      
      for (l in 1:nfec) {
        ft.fec.data1[l] <- ((Nobs_juv2[l]-(py_j*Njuv2[l]))^2)/(py_j*Njuv2[l])
        ft.fec.pred1[l] <- ((Nobs_juv.pred[l]-(py_j*Njuv2[l]))^2)/(py_j*Njuv2[l])
      }#l
      
      ft.fec.data2 <- sum(ft.fec.data1[])
      ft.fec.pred2 <- sum(ft.fec.pred1[])
      
      pvalue.fec <- step(ft.fec.pred2-ft.fec.data2)
      
      # Part 5.2: Capture History
      
      for (i in 1:nind) {
        ### The next two lines are adjustments for code to run
        ### They don't affect the calculation of Freeman-Tukey statistic
        p[i,first[i],first_sub[i]] <- 1
        start[i] <- ifelse(first_sub[i]==nsub, nsub, first_sub[i]+1)
      
        ft1.data[i,first[i]] <- ifelse(first_sub[i]==nsub, 
                                       0, 
                                       (sqrt(sum(y[i,first[i],start[i]:nsub])) - 
                                        sqrt(sum(p[i,first[i],start[i]:nsub])))^2)
      
        ft1.pred[i,first[i]] <- ifelse(first_sub[i]==nsub, 
                                       0, 
                                       (sqrt(sum(y.pred[i,first[i],start[i]:nsub])) - 
                                        sqrt(sum(p[i,first[i],start[i]:nsub])))^2)
      
        for (t in (first[i]+1):nyear) {
          ft1.data[i,t] <- (sqrt(sum(y[i,t,])) - sqrt(z[i,t]*sum(p[i,t,])))^2
          ft1.pred[i,t] <- (sqrt(sum(y.pred[i,t,])) - sqrt(z[i,t]*sum(p[i,t,])))^2
        }#t
      
        ft2.data[i] <- sum(ft1.data[i,first[i]:nyear])
      
        ft2.pred[i] <- sum(ft1.pred[i,first[i]:nyear])
      }#i
      
      ft3.data <- sum(ft2.data[])
      ft3.pred <- sum(ft2.pred[])
      
      pvalue.ch <- step(ft3.pred-ft3.data)
      
    }", fill = TRUE)
sink()
}
```


### 2. Density independent model

```{r, message=FALSE}
{
sink("di_model_sim.jags")
cat(
  "model {
    
    # Part 1: Priors
      
      # Part 1.1: Priors for survival model
      
      survival_ad ~ dunif(0,1)
      survival_juv ~ dunif(0,1)
      alpha[2] <- logit(survival_ad)
      alpha[1] <- logit(survival_juv)
      sigma_s ~ dt(0,pow(10,-2),1)T(0,)
      tau_s <- 1/(sigma_s^2)
      
      for (k in 1:npop) {
        for (t in first_pop[k]:(nyear-1)) {
          epsilon[k,t] ~ dnorm(0, tau_s)
        }
      }
      
      # Part 1.2: Priors for capture probability model
      
      gamma[1] ~ dt(0,pow(10,-2),1)
      gamma[2] ~ dt(0,pow(10,-2),1)
      
      # Part 1.3: Priors for fecundity model
      
      fecundity ~ dnorm(0,pow(10,-2))T(0,)
      theta <- log(fecundity)
      sigma_f ~ dt(0,pow(5,-2),1)T(0,)
      tau_f <- 1/(sigma_f^2)
      
      for (l in 1:nfec) {
        omega[l] ~ dnorm(0,tau_f)
      }

    
    # Part 2: Robust design mark-recapture model
      
    for (i in 1:nind) {
        
      # Part 2.1: Survival process
        
      z[i,first[i]] ~ dbern(1)
      
      for (t in first[i]:(nyear-1)) {
        phi[i,t] <- ilogit(alpha[stage[i,t]] + epsilon[pop[i],t])
        z[i,t+1] ~ dbern(phi[i,t]*z[i,t])
      }#t
      
      # Part 2.2: Observation
      
      y[i,first[i],first_sub[i]] ~ dbern(1)
      y.pred[i,first[i],first_sub[i]] ~ dbern(1)
      
      for(h in (first_sub[i]+1):nsub) {
        p[i,first[i],h] <- ilogit(gamma[stage[i,first[i]]])
        y[i,first[i],h] ~ dbern(z[i,first[i]]*p[i,first[i],h])
        y.pred[i,first[i],h] ~ dbern(z[i,first[i]]*p[i,first[i],h])
      }#h
      
      for (t in first[i]:(nyear-1)) {
        for(h in 1:nsub) {
          p[i,t+1,h] <- ilogit(gamma[stage[i,t+1]])
          y[i,t+1,h] ~ dbern(z[i,t+1]*p[i,t+1,h])
          y.pred[i,t+1,h] ~ dbern(z[i,t+1]*p[i,t+1,h])
        }#h
      }#t
    }#i
      
    
    # Part 3: Population size and Density Estimation
      
      # Part 3.1: Probability of capture per population and year
      
      py_a <- 1 - pow(1-ilogit(gamma[1]),4)
      py_j <- 1 - pow(1-ilogit(gamma[2]),4)
      
      # Part 3.2: Density estimation
      for (k in 1:npop) {
        for (t in first_pop[k]:nyear) {
          Nad[k,t] <- (Nobs_ad[k,t]/py_a) + (1-py_a)/py_a
          Njuv1[k,t] <- (Nobs_juv1[k,t]/py_j) + (1-py_j)/py_j
        }#t
      }#k
      
    
    # Part 4: Fecundity model
      
      for (l in 1:nfec) {
        year[l] <- year.index[l]-1
        Njuv2[l] <- exp(log(Nad[pop.index[l],year.index[l]]) + theta + omega[l])
        Nobs_juv2[l] ~ dpois(max(py_j*Njuv2[l], 0.00000001))
        Nobs_juv.pred[l] ~ dpois(max(py_j*Njuv2[l], 0.00000001))
      }#l
      
    
    # Part 5: Model Checking
      
      # Part 5.1: Fecundity
      
      for (l in 1:nfec) {
        ft.fec.data1[l] <- ((Nobs_juv2[l]-(py_j*Njuv2[l]))^2)/(py_j*Njuv2[l])
        ft.fec.pred1[l] <- ((Nobs_juv.pred[l]-(py_j*Njuv2[l]))^2)/(py_j*Njuv2[l])
      }#l
      
      ft.fec.data2 <- sum(ft.fec.data1[])
      ft.fec.pred2 <- sum(ft.fec.pred1[])
      
      pvalue.fec <- step(ft.fec.pred2-ft.fec.data2)
      
      # Part 5.2: Capture History
      
      for (i in 1:nind) {
        ### The next two lines are adjustments for code to run
        ### They don't affect the calculation of Freeman-Tukey statistic
        p[i,first[i],first_sub[i]] <- 1
        start[i] <- ifelse(first_sub[i]==nsub, nsub, first_sub[i]+1)
      
        ft1.data[i,first[i]] <- ifelse(first_sub[i]==nsub, 
                                       0, 
                                       (sqrt(sum(y[i,first[i],start[i]:nsub])) - 
                                        sqrt(sum(p[i,first[i],start[i]:nsub])))^2)
      
        ft1.pred[i,first[i]] <- ifelse(first_sub[i]==nsub, 
                                       0, 
                                       (sqrt(sum(y.pred[i,first[i],start[i]:nsub])) - 
                                        sqrt(sum(p[i,first[i],start[i]:nsub])))^2)
      
        for (t in (first[i]+1):nyear) {
          ft1.data[i,t] <- (sqrt(sum(y[i,t,])) - sqrt(z[i,t]*sum(p[i,t,])))^2
          ft1.pred[i,t] <- (sqrt(sum(y.pred[i,t,])) - sqrt(z[i,t]*sum(p[i,t,])))^2
        }#t
      
        ft2.data[i] <- sum(ft1.data[i,first[i]:nyear])
      
        ft2.pred[i] <- sum(ft1.pred[i,first[i]:nyear])
      }#i
      
      ft3.data <- sum(ft2.data[])
      ft3.pred <- sum(ft2.pred[])
      
      pvalue.ch <- step(ft3.pred-ft3.data)
      
    }", fill = TRUE)
sink()
}
```